---
title: "Exploratory analysis of storm events"
author: "I Carter (Coursera)"
output: 
  html_document:
    keep_md: true
    toc: true
---

## Synopsis

This file contains my Exploratory Analysis for the second peer reviewed project.

## Exploratory Analysis

### Read in dataset

```{r echo = TRUE}
bz <- bzfile("repdata_data_StormData.csv.bz2", "r")
data <- read.csv(bz)
close(bz)
```

### Take a first look at dataset

```{r echo = TRUE}
dim(data)       ## 902297 records, 37 columns
str(data)
colnames(data)
```

----

### Looking at events per state

#### Mystery States

```{r echo = TRUE}
str(data$STATE)                                 ## 72 levels???
temp <- levels(data$STATE) %in% state.abb       ## which are not in state.abb?
levels(data$STATE)[!temp]                       ## 22 mystery state codes...
```

Mystery states turned out to be places like Guam, as a look at `BGN_LOCATI`, `STATEOFFIC`, or `COUNTYNAME` shows (`head(data[data$STATE=="GU",])`).

#### How many events for the non-standard state codes?

```{r echo = TRUE}
states <- summary(data$STATE)
states[levels(data$STATE)[!temp]]     ## look at the 22 mystery states
other <- sum(states[levels(data$STATE)[!temp]])
other                                 ## 19111
other/sum(states) * 100               ##
```

#### Which State has had the most events?

```{r echo = TRUE}
max(states)                           ## 83728
which.max(states)                     ## Texas (TX, 63)
sum(states["TX"])/sum(states) * 100   ## 
```

----

### Looking at dates

#### Do all the dates "play nicely"?

I.e. does `POSIXlt()` work OK with all the dates?

```{r echo = TRUE}
sum(is.na(as.POSIXlt(levels(data$BGN_DATE), format="%m/%d/%Y")))        ## 0
sum(is.na(as.POSIXlt(levels(data$END_DATE), format="%m/%d/%Y")))        ## 1
```

#### A Closer Look at End Dates

```{r echo = TRUE}
str(data$END_DATE)      ## some of the end dates are missing values (i.e. "")
sum(data$END_DATE == "")                                ## 243411 missing values
sum(data$END_DATE == "")/length(data$END_DATE) * 100    ## about 27%
```

`END_DATE` data are very incomplete.

#### Range of dates

```{r echo = TRUE}
min(as.POSIXlt(levels(data$BGN_DATE), format="%m/%d/%Y"))               ## "1950-01-03 GMT"
max(as.POSIXlt(levels(data$BGN_DATE), format="%m/%d/%Y"))               ## "2011-11-30 GMT"
min(as.POSIXlt(levels(data$END_DATE), format="%m/%d/%Y"), na.rm=TRUE)   ## "1986-04-10 BST"
max(as.POSIXlt(levels(data$END_DATE), format="%m/%d/%Y"), na.rm=TRUE)   ## "2011-11-30 GMT"
```

So begin dates go from January 1950 to the end of November 2011, whereas end dates only start in April 1986.

----

### Numeric values of `PROPDMGEXP`

We'll have a look at some of the records where `PROPDMGEXP` is 4, 5, 6, 7, or 8. If `PROPDMGEXP` is indeed an exponent these events would be pretty costly.

```{r echo = TRUE}
prop4 <- data[data$PROPDMGEXP == 4,]    ## 4 records
prop5 <- data[data$PROPDMGEXP == 5,]    ## 28 records
prop6 <- data[data$PROPDMGEXP == 6,]    ## 4 records
prop7 <- data[data$PROPDMGEXP == 7,]    ## 5 records
prop8 <- data[data$PROPDMGEXP == 8,]    ## 1 record
as.character(prop4$REMARKS)
as.character(prop5$REMARKS)
as.character(prop6$REMARKS)
as.character(prop7$REMARKS)
as.character(prop8$REMARKS)
```

#### Strange data in `prop5`

We see that, for example, `as.character(prop5$REMARKS)[11]` says that the total damage (from lightning striking a house) was estimated to be €3500, but while `prop5$PROPDMG` is indeed 3, the exponent should be 3 also, not 5, so this is 2 orders of magnitude wrong.

Other examples of damage estimates that couldn't possibly be right (from `prop5`) include 22 ("Several homes and several hundered vehicle were damaged. ") and 24, where the awning of a shopping center and a large section of an hotel's roof was torn off. These both have `PROPDMG` set to 0, obviously wrong.

"It took more than 100 firefighters from 19 communities many hours to extinguish a fire started by lightning which destroyed a large barn at a dairy farm in Westport." However `PROPDMG` here is only 0.2, with `PROPDMGEXP == 5`. I simply cannot believe the damage described is only $20,000.

We also see that, while we have removed the empty comments, some of the remarks are `"  "` (2 spaces), which is not helpful.

#### Strange data in `prop7` and `prop8`

In the first event the police station roof and radio tower were bady damaged, and some trees were uprooted. But with `PROPDMG` of 14 and `PROPDMGEXP` of 7, I can't believe that this damage cost $140 million.

The one record ("Several homes and several hundered vehicle were damaged.") with `PROPDMGEXP == 8` has `PROPDMG` of 0.

#### More exploration of `PROPDMGEXP`

```{r echo = TRUE}
## Get all the records with PROPDMGEXP of 0 to 8
prop_num <- data[data$PROPDMGEXP %in% 0:8,]
temp <- summary(prop_num$EVTYPE)
temp[temp!=0]                           ## print out a table of event types
dim(temp)[1] / dim(prop_num)[1] * 100
## We look at the percentage of "THUNDERSTORM WINDS" in the entire dataset
temp <- data[data$EVTYPE=="THUNDERSTORM WINDS",]
dim(temp)[1]/dim(data)[1] * 100         ## only 2.3%
```

we see `"THUNDERSTORM WINDS"` are highly over-represented (64%) in the data that has a numeric `PROPDMGEXP`, compared to 2.3% in the full dataset.

Maybe we would find more patterns if we looked for them...

#### Conclusion

We conclude that the numeric values of `PROPDMGEXP` don't seem to have a consistant meaning.

----

### Records where `PROPDMG == 0`

#### How many events cause no property damage or no crop damage?

```{r echo = TRUE}
prop0 <- data[data$PROPDMG == 0,]
dim(prop0)[1]/dim(data)[1] * 100        ## 73.5%
crop0 <- data[data$CROPDMG == 0,]
dim(crop0)[1]/dim(data)[1] * 100        ## 97.5%
```

```{r echo = TRUE}
temp <- as.character(prop0$REMARKS)
temp <- temp[temp!=""]
temp[1:12]
```

Well, this means that we can't assume that `PROPDMG == 0` means no property damage. For example one of the `REMARKS` where `PROPDMG == 0` explicitly states "Damage to homes and businesses was widespread." Presumably the data for crop damage is similarly flaky.

But we can only deal with the data we have. We just need to bear in mind that it is not fully accurate or complete.

----

### `usmap`

```{r echo = TRUE}
library(usmap)
library(ggplot2)
```

#### Everything goes wrong in Texas...

```{r echo = TRUE}
temp <- summary(data$STATE)
temp <- temp[state.abb]
temp <- data.frame(state=names(temp), events=as.numeric(temp))
plot_usmap(data = temp, values="events", color = "red") +
    scale_fill_continuous(low="green", high="red",
        name = "# Events", label = scales::comma) + 
    theme(legend.position = "right")
```

#### One county seems more dangerous than the others in Arkansas...

```{r echo = TRUE}
arkansas <- data[data$STATE=="AR",]
## ar_counties <- levels(as.factor(arkansas$COUNTY))
ar_counties <- format(arkansas$COUNTY, width=3)
ar_counties <- sapply(ar_counties, function(x) {gsub(" ", "0", x)} )
arkansas$COUNTY <- sapply(ar_counties, function(x) {paste("05", x, sep="")})
test <- summary(as.factor(arkansas$COUNTY))
test <- data.frame(fips=names(test), events=as.numeric(test))
plot_usmap(data = test, values="events", color = "red",
    include=test$fips)
```

#### Tornados in Texas

```{r echo = TRUE}
tornado <- data[grep("TORNADO", data$EVTYPE),]
dim(tornado)[1] / dim(data)[1] * 100                    ## 6.73% nationally
tornado_tx <- subset(tornado, STATE=="TX")
dim(tornado_tx)[1] / sum(data$STATE=="TX") * 100        ## 9.90% in TX
tx_counties <- format(tornado_tx$COUNTY, width=3)
tx_counties <- sapply(tx_counties, function(x) {gsub(" ", "0", x)} )
tornado_tx$COUNTY <- sapply(tx_counties, function(x) {paste("48", x, sep="")})
test <- summary(as.factor(tornado_tx$COUNTY))
test <- data.frame(fips=names(test), events=as.numeric(test))
plot_usmap("counties", data = test, values="events", color = "red",
    include = "TX")
```

----

This document was processed on: `r Sys.Date()`.
