# Analysis of U.S. National Oceanic and Atmospheric Administration's (NOAA) Storm Database

## Data Processing

### Read in dataset

```{r echo = TRUE}
bz <- bzfile("repdata_data_StormData.csv.bz2", "r")
data <- read.csv(bz)
close(bz)
```

### Take a first look at dataset

```{r echo = TRUE}
dim(data)       ## 902297 records, 37 columns
str(data)
colnames(data)
```

 [1] "STATE__"    "BGN_DATE"   "BGN_TIME"   "TIME_ZONE"  "COUNTY"    
 [6] "COUNTYNAME" "STATE"      "EVTYPE"     "BGN_RANGE"  "BGN_AZI"   
[11] "BGN_LOCATI" "END_DATE"   "END_TIME"   "COUNTY_END" "COUNTYENDN"
[16] "END_RANGE"  "END_AZI"    "END_LOCATI" "LENGTH"     "WIDTH"     
[21] "F"          "MAG"        "FATALITIES" "INJURIES"   "PROPDMG"   
[26] "PROPDMGEXP" "CROPDMG"    "CROPDMGEXP" "WFO"        "STATEOFFIC"
[31] "ZONENAMES"  "LATITUDE"   "LONGITUDE"  "LATITUDE_E" "LONGITUDE_"
[36] "REMARKS"    "REFNUM"    

### How many fatalities and injuries?

```{r echo = TRUE}
sum(data$FATALITIES)    ## 15145
sum(data$INJURIES)      ## 140528
```

### Looking at events per state

#### Mystery States

```{r echo = TRUE}
str(data$STATE)                                 ## 72 levels???
temp <- levels(data$STATE) %in% state.abb       ## which are not in state.abb?
levels(data$STATE)[!temp]                       ## 22 mystery state codes...
```
#### How many events for the mystery states?

```{r echo = TRUE}
states <- summary(data$STATE)
states[levels(data$STATE)[!temp]]       ## look at the 22 mystery states
sum(states[levels(data$STATE)[!temp]])  ## 19111
```

### Which State has had the most events?

```{r echo = TRUE}
max(summary(data$STATE))        ## 83728
which.max(summary(data$STATE))  ## Texas (TX, 63)
```

### Looking at dates

#### Do all the dates "play nicely"?

I.e. does `POSIXlt()` work OK with all the dates?

```{r echo = TRUE}
sum(is.na(as.POSIXlt(levels(data$BGN_DATE), format="%m/%d/%Y")))        ## 0
sum(is.na(as.POSIXlt(levels(data$END_DATE), format="%m/%d/%Y")))        ## 1
```

#### A Closer Look at End Dates

```{r echo = TRUE}
str(data$END_DATE)      ## some of the end dates are missing values (i.e. "")
sum(data$END_DATE == "")                                ## 243411 missing values
sum(data$END_DATE == "")/length(data$END_DATE) * 100    ## about 27%
```

```{r echo = TRUE}
```

```{r echo = TRUE}
```

```{r echo = TRUE}
```

## Results

The questions that need answers are:

### Most Harmful EVTYPEs

Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

### Economic Consequences

Across the United States, which types of events have the greatest economic consequences?
